---
params:
  title: 'Covid19 Data Explorations'
  subtitle: 'USA deaths data'
  authors: 'Ben Anderson (`@dataknut`)'
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
always_allow_html: yes
bibliography: '`r path.expand("~/bibliography.bib")`'
---

```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages ----
rmdLibs <- c("ggplot2",
             "here",
            "kableExtra",
            "lubridate",
            "plotly",
            "skimr")

dkUtils::loadLibraries(rmdLibs)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)



# Parameters ----
myParams <- list()
myParams$dPath <- paste0(here::here(), "/data")

# Log compile time:
myParams$startTime <- proc.time()

# set values for annotations
myParams$myAlpha <- 0.1
myParams$vLineAlpha <- 0.4
myParams$vLineCol <- "#0072B2" # http://www.cookbook-r.com/Graphs/Colors_(ggplot2)/#a-colorblind-friendly-palette
myParams$myTextSize <- 4
myParams$rectAlpha <- 0.05

myParams$ldStart <- as.Date("2020-03-23") # not true but...
myParams$ldStart <- as.Date("2020-04-23") # not true but...

addLockdownDate <- function(p, start, finish){
  # assumes p has x = obsDate
  p <- p + annotate("text", x = lubridate::date(start),
             y = yMax * 0.4, angle = 10,size = myTextSize,
             label = "UK covid lockdown", hjust = 0)
  p <- p + annotate("rect", xmin = lubridate::as_date(start),
             xmax = lubridate::as_date(finish), 
             ymin = yMin, ymax = yMax, 
             alpha = rectAlpha, 
             fill = vLineCol, 
             colour = vLineCol)
  return(p)
}

```

# About

## Contributions

Please note that authorship is alphabetical. Contributions are listed below - see [github](https://github.com/dataknut/covid19/commits/master) for details and who to blame for what :-).

 * Ben Anderson (b.anderson@soton.ac.uk `@dataknut`)
 * 

## Code

 * https://github.com/dataknut/covid19/commits/master
 
## Making comments
Free and easy to do. Raise an issue at:

 * https://github.com/dataknut/covid19/issues

# Intro

Exploring covid 19 and related data.

This tweet:

 * https://twitter.com/dweekly/status/1244419047510306817 and the paper it linked to: https://www.dropbox.com/s/2y2gvh2q7axdzz2/COVID-19_UnintendedConsequences.pdf?dl=0

Suggested something strange was happening with overall USA deaths in early 2020.:

![USA deaths plot](https://pbs.twimg.com/media/EUUQVGZUEAEpMsT?format=jpg&name=small)

This very brief analysis attempts to re-create some of their results from the same data.

# Data

Sourced from the downloads button of https://gis.cdc.gov/grasp/fluview/mortality.html

> "The National Center for Health Statistics (NCHS) collects and disseminates the Nation's official vital statistics. NCHS collects death certificate data from state vital statistics offices for virtually all deaths occurring in the United States. Pneumonia and influenza (P&I) deaths are identified based on ICD-10 multiple cause of death codes."

```{r loadData}
dt <- data.table::fread(paste0(myParams$dPath, "/USA-all-data.csv"))

t <- table(dt$SEASON)
kableExtra::kable(t, caption = "Seasons used") %>%
  kable_styling()
```

Note that the 2020 data is incomplete.

Before we go further we'll convert week numbers (in the data) to standardised dates. This will mean some of the dates for the weeks are not _exactly_ right but it might make for easier plotting.

```{r tidyData}
# why can't they just be published with dates??

dt[, date := ifelse(SEASON == "2015-16" & WEEK > 39, as.Date("2015-01-01") + lubridate::weeks(WEEK),
                    NA)]
dt[, date := ifelse(SEASON == "2015-16" & WEEK < 40, as.Date("2016-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2016-17" & WEEK > 39, as.Date("2016-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2016-17" & WEEK < 40, as.Date("2017-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2017-18" & WEEK > 39, as.Date("2017-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2017-18" & WEEK < 40, as.Date("2018-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2018-19" & WEEK > 39, as.Date("2018-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2018-19" & WEEK < 40, as.Date("2019-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2019-20" & WEEK > 39, as.Date("2019-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := ifelse(SEASON == "2019-20" & WEEK < 40, as.Date("2020-01-01") + lubridate::weeks(WEEK),
                    date)]
dt[, date := lubridate::as_date(date)]

# check coding
# head(dt[WEEK == 1])
t <- head(dt[WEEK == 1 | WEEK == 52, .(SEASON, WEEK, date)])

kableExtra::kable(t, captopn = "Check we coded the weeks correctly") %>%
  kable_styling()

# also clean up data where needed
dt[, totDeaths := as.numeric(gsub(",", "",`TOTAL DEATHS`))] 
dt[, pneuDeaths := as.numeric(gsub(",", "",`NUM PNEUMONIA DEATHS`))]
dt[, fluDeaths := as.numeric(gsub(",", "",`NUM INFLUENZA DEATHS`))]

dt$V13 <- NULL
dt$V14 <- NULL
dt$V15 <- NULL
dt$V16 <- NULL
dt$V17 <- NULL
dt$V18 <- NULL

skimr::skim(dt)
tdt <- dt[`PERCENT COMPLETE` != "> 100%"]
kableExtra::kable(table(tdt$date, tdt$`PERCENT COMPLETE`), caption = "Incomplete data...") %>%
  kable_styling()

dt <- dt[`PERCENT COMPLETE` == "> 100%"]
```

We remove the incomplete data from further analysis.

Check on total number of deaths per season.

```{r sumDeaths}
t <- dt[, .(Total = dkUtils::tidyNum(sum(totDeaths)),
            Pneumonia = dkUtils::tidyNum(sum(pneuDeaths)),
            Influenza = dkUtils::tidyNum(sum(fluDeaths))),
        keyby = .(SEASON)]
kableExtra::kable(t, caption = "Sum of deaths by season - do these look sensible?", ) %>%
  kable_styling()
```

# All deaths (all ages)

Figure \@ref(fig:plotAllDeaths) is the one causing the interest. Note that we have set the y axis to start at zero to avoid over-emphasising the slope. Week 10 is roughly:

```{r checkWeeks}
dt[WEEK == 10, .(SEASON, WEEK, date)]
```

> Note that there are potential issues with the data. See Section \@ref(conclusions).

```{r plotAllDeaths, fig.cap="USA recorded deaths by flu season (all deaths, all ages)"}
week10Text <- paste0("Week 10: ", dt[WEEK == 10 & SEASON == "2019-20", (date)])
addWeek10Label <- function(p){
  p <- p + annotate("text", x = 10.5,
             y = yMax * 0.4, angle = 10,size = 3,
             label = week10Text, hjust = 0)
  p <- p + geom_vline(xintercept = 10, colour = "#CC79A7", alpha = myParams$vLineAlpha)
  return(p)
}

p <- ggplot2::ggplot(dt, aes(y = totDeaths, x = WEEK, colour = SEASON, group = SEASON)) +
  geom_point() +
  ylim(0,NA) +
  labs(x = "Week",
       y = "Total deaths",
       caption = "Plot by @dataknut using data from https://gis.cdc.gov/grasp/fluview/mortality.html\nIncomplete data excluded")
yMax <- max(dt$totDeaths)
p <- addWeek10Label(p)
p
```

Figure \@ref(fig:plotlyAllDeaths): interactive version (hover the mouse over a dot). 

```{r plotlyAllDeaths, fig.cap="USA recorded deaths by flu season (all deaths)"}
plotly::ggplotly(p)
```

# Deaths from pneumonia (all ages)

Figure \@ref(fig:plotPneuDeaths) is pneumonia deaths. They are trending downwards too.

```{r plotPneuDeaths, fig.cap="USA recorded Pneumonia deaths by flu season, all ages"}
p <- ggplot2::ggplot(dt, aes(y = pneuDeaths, x = WEEK, colour = SEASON, group = SEASON)) +
  geom_point() +
  ylim(0,NA)+
  labs(x = "Week",
       y = "Pneumonia deaths",
       caption = "Plot by @dataknut using data from https://gis.cdc.gov/grasp/fluview/mortality.html")
yMax <- max(dt$pneuDeaths)
p <- addWeek10Label(p)
p
```

The 'missing' deaths here don't seem to account for the overall total deaths reduction. It seems that pneumonia deaths are not being re-classed as covid since USA covid deaths are not recorded as starting until [after March 10th](https://virusncov.com/covid-statistics/usa). Although it is _possible_ that earlier covid deaths were not recorded as covid _or_ pneuomina (_or_ influenza).

# Deaths from influenza (all ages)

Figure \@ref(fig:plotFluDeaths) is influenza deaths. 

```{r plotFluDeaths, fig.cap="USA recorded Influenza deaths by flu season, all ages"}
p <- ggplot2::ggplot(dt, aes(y = fluDeaths, x = WEEK, colour = SEASON, group = SEASON)) +
  geom_point() +
  ylim(0,NA)+
  labs(x = "Week",
       y = "Influenza deaths",
       caption = "Plot by @dataknut using data from https://gis.cdc.gov/grasp/fluview/mortality.html")
yMax <- max(dt$fluDeaths)
p <- addWeek10Label(p)
p
```

Yep, 2017-2018 was a [bad flu year in the USA](https://www.cdc.gov/flu/about/season/flu-season-2017-2018.htm)...

# Conclusions {#conclusions}

> The decline in 2020 'all deaths' appears real provided the 2020 data is complete. For more on this question, see:

 * https://twitter.com/SidSanghi/status/1244344193230819328

We can speculate about why and there are many suggestions in the thread sparked by the tweet:

 * https://twitter.com/dweekly/status/1244419047510306817

# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - myParams$startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used:

 * data.table - [@data.table]
 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * lubridate - [@lubridate]
 * plotly - [@plotly]
 * skimr - [@skimr]
            
# References
