---
params:
  title: ''
  subtitle: ''
  authors: ''
title: '`r params$title`'
subtitle: '`r params$subtitle`'
author: '`r params$authors`'
date: 'Last run at: `r Sys.time()`'
output:
  bookdown::html_document2:
    fig_caption: yes
    code_folding: hide
    self_contained: true # set to false if you want plots seperately for easy import to word https://bookdown.org/yihui/rmarkdown/html-document.html#document-dependencies
    number_sections: yes
    toc: yes
    toc_depth: 2
    toc_float: TRUE
bibliography: '`r path.expand("~/bibliography.bib")`'
---
  
```{r knitrSetUp, include=FALSE}

# Set knitr options
knitr::opts_chunk$set(echo = FALSE,          # echo code so reader can see what is happening
                      warning = FALSE,
                      message = FALSE,
                      fig_caption = TRUE,
                      fig_height = 6,        # default, make it bigger to stretch vertical axis
                      fig_width = 8,
                      fig_width = 8,         # full width
                      tidy = TRUE)           # tidy up code in case echo = TRUE

```


```{r codeSetup, include=FALSE}

# Load Packages needed only in Rmd ----
library(kableExtra)

# Adjust knitr options if required
knitr::opts_chunk$set(echo = TRUE)

# Local parameters ----

# Log compile time:
myParams$startTime <- proc.time()

```

# About

## Code

Source:

 * https://github.com/dataknut/covid19

History:

 * https://github.com/dataknut/covid19/commits/master

## Comments and feedback

If you wish to comment please open an issue:
 
 * https://github.com/dataknut/covid19/issues

## Citation

If you wish to refer to any of the material from this report please cite as:

 * `r params$authors`, (`r format(Sys.time(), format = "%Y")`) `r params$title`: `r params$subtitle` , [Sustainable Energy Research Group](http://www.energy.soton.ac.uk), University of Southampton: Southampton, UK.

Report circulation:

 * Public
 
This work is (c) `r format(Sys.time(), format = "%Y")` the University of Southampton.
 
# Introduction

An attempt to learn text mining using Boris Johnson's televised address on May 10th 2020.

Data: https://www.gov.uk/government/speeches/pm-address-to-the-nation-on-coronavirus-10-may-2020

Methods draw very heavily on https://www.tidytextmining.com ([Text Mining with R: A Tidy Approach](http://shop.oreilly.com/product/0636920067153.do), Julia Silge and David Robinson, 2020-03-07)

# The data

A [relatively short speech/address](https://www.gov.uk/government/speeches/pm-address-to-the-nation-on-coronavirus-10-may-2020) of `r nrow(df)` sentences outlining the framework for easing UK (actually English) lockdown.


```{r dataDesc}
nWords <- nrow(tidy_df)
nWordsReduced <- nrow(reduced_df)

kableExtra::kable(head(df), caption = "First 6 sentences") %>%
  kable_styling()
```

The address used a total of `r nWords` different words of which `r nWordsReduced` were `non-stopping` words (if, and, the etc)

# Word frequency by postive/negative sentiment

We use the [`bing` sentiment data](https://www.cs.uic.edu/~liub/FBS/sentiment-analysis.html) to assign `sentiment` (positive vs negative) to each non-stopping word used and calculate the frequency.

Figure \@ref(fig:bingWordCountsPlot) shows the frequency of these words by sentiment.

```{r bingWordCountsPlot, fig.cap="Frequency of words by sentiment"}
bing_word_countsPlot
```

Figure \@ref(fig:wordCloudSentimentPlot) shows a word cloud of these words by sentiment.

```{r wordCloudSentimentPlot, fig.cap="Wordcloud of words by sentiment"}
# for some reson we can't capture this as a plot in the R script?
wc <- reduced_df %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  reshape2::acast(word ~ sentiment, value.var = "n", fill = 0)

wordcloud::comparison.cloud(wc, colors = c("gray80", "gray20"),
                   max.words = 100)
```

# Frequency of sentiment by line

Figure \@ref(fig:sentimentByLinePLot) shows the number of positive/negative words by line (sentence) number. As we can see there are positive words at the very start followed by negative words, then a mix and then a positive finale.

```{r sentimentByLinePLot, fig.cap="Frequency of entiment by line number"}
sentimentByLinePLot
```

# Runtime

Report generated using [knitr](https://cran.r-project.org/package=knitr) in [RStudio](http://www.rstudio.com) with `r R.version.string` running on `r R.version$platform` (`r Sys.info()[3]`).

```{r check runtime}
t <- proc.time() - myParams$startTime

elapsed <- t[[3]]
```

Analysis completed in `r elapsed` seconds ( `r round(elapsed/60,2)` minutes).

R packages used in this report:

 * ggplot2 - [@ggplot2]
 * here - [@here]
 * kableExtra - [@kableExtra]
 * readr = [@readr]
 * reshape2 - [@reshape2]
 * tidytext - [@tidytext]
 * tidyverse - [@tidyverse]
 * wordcloud - [@wordcloud]
 
# References
